---
title: "p8105_hw2_lz3015"
output: github_document
date: "2025-09-30"
---
## Loading packages
```{r}
library(tidyverse)
library(dplyr)
```

## Problem 1
cleaning pols-month
```{r}
pols_df = 
  read.csv(
    "./fivethirtyeight_data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day"), sep = "-") |> 
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    month = month.name[month], 
    president= if_else(prez_gop == 1, "gop", "dem")
  ) |> 
  select(-prez_dem, -prez_gop, -day) #drops these variables
```
cleaning snp
```{r}
snp_df = 
  read.csv(
    "./fivethirtyeight_data/snp.csv") |> 
  janitor::clean_names() |> 
  separate(date, into = c("month", "day", "year"), sep = "/") |> 
  mutate(
    year = as.integer(year),
    year = if_else(year > 20, 1900 + year, 2000 + year),
    month = as.integer(month),
    month = month.name[month], 
  ) |> 
  select(-day) |> 
  arrange(year)
```
cleaning and tidying unemployment
```{r}
unemployment_df = 
    read.csv(
      "./fivethirtyeight_data/unemployment.csv") |>
  rename(year = Year) |> # removing capital y for consistency
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment") |> 
  mutate(
    month = match(month, month.abb),
    month = month.name[month]
  )
    
```
Joining the three datasets
```{r}
# join with unemployment data
join_df = pols_df |>
  left_join(snp_df, by = c("year", "month")) |>
  left_join(unemployment_df, by = c("year", "month"))
```

## Data description:
The `pols_df` data contains 822 observations related to the number of national politicians who are democratic or republican at any given time.
This data set contains information as early as January 1947 and goes all the way to June of 2015 and there are `r ncol(pols_df)` variables

The `snp_df` data contains 787 observations related to Standard & Poor’s stock market index (S&P) represented by the variable "close" (the closing 
values of the S&P stock index on the associated date),often used as a representative measure of stock market as a whole. This data was also tracked
over time starting in February of 2001 and ending in 2012 and there are `r ncol(snp_df)` variables

The `unemployment_df` data contains 816 observations detailing the unemployment rate between January 1948 and December 2015 and there are
`r ncol(unemployment_df)` variables

The `join_df` data is the combination of the three data sets described above with `r nrow(join_df)` observations and `r ncol(join_df)` variables



## Problem 2
```{r}
library(readxl)
```
Mr. Trash Wheel
```{r}
mr_trash_df = read_excel("trashwheel_data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    skip = 1) |> 
  janitor::clean_names() |> 
  select(-starts_with("x")) |>  # drops empty columns
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel",
    year = as.integer(year) # Need to standardize data type for year for future joins
  )
```

```{r}
prof_trash_df = read_excel("trashwheel_data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    skip = 1) |> 
  janitor::clean_names() |> 
  select(-starts_with("x")) |>  # drops empty columns
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Professor Trash Wheel",
    year = as.integer(year) # Need to standardize data type for year for future joins
  )
```

```{r}
gwynnda_trash_df = read_excel("trashwheel_data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    skip = 1) |> 
  janitor::clean_names() |> 
  select(-starts_with("x")) |>  # drops empty columns
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Gwynns Falls Trash Wheel" ,
    year = as.integer(year) # Need to standardize data type for year for future joins
  )
```

```{r}
all_trash_df = 
  bind_rows(
    mr_trash_df,
    prof_trash_df,
    gwynnda_trash_df
)
```

The `mr_trash_df` data contains `r nrow(mr_trash_df)` observations and `r ncol(mr_trash_df)` variables. 
The `prof_trash_df` data contains `r nrow(prof_trash_df)` observations and `r ncol(prof_trash_df)` variables. 
The `gwynnda_trash_df` data contains `r nrow(gwynnda_trash_df)` observations and `r ncol(gwynnda_trash_df)` variables. 
`weight_tons` and `volume_cubic_yards` are among the main descriptive variables that illustrate the magnitude of the trash collected.
The average weight collected on a given day in tons for each wheel is:
Mr. Trash Wheel = `r round(mean(mr_trash_df$weight_tons, na.rm = TRUE), 1)`;
Professor Trash Wheel = `r round(mean(prof_trash_df$weight_tons, na.rm = TRUE), 1)`;
Gwynnda Trash Wheel = `r round(mean(gwynnda_trash_df$weight_tons, na.rm = TRUE), 1)`;
The overall average `weight_tons` from the combined `all_trash_df` is `r round(mean(all_trash_df$weight_tons, na.rm = TRUE), 1)`
The recorded data spans from `r min(all_trash_df$year, na.rm = TRUE)` to `r max(all_trash_df$year, na.rm = TRUE)`

## Problem 3
```{r}
zip_df = 
  read.csv (
  "./zillow_data/Zip Codes.csv") |> 
  janitor::clean_names() |> 
  rename(
    borough = county
  )

#Renaming borough names 
zip_df = zip_df |> 
  mutate(
    borough = recode(
    borough,
    "Richmond" = "Staten Island",
    "New York" = "Manhattan",
    "Kings" = "Brooklyn"
))
```
View df to make sure columns have proper labels and boroughs were recoded
```{r}
view(zip_df)
```
After looking at view, we should check for zip code duplicates
```{r}
zip_df |> 
  group_by(zip_code) |> 
  filter(n() > 1) |> 
  ungroup()
```
Remove duplicates:
```{r}
zip_df_unique = zip_df |> 
  distinct(zip_code, .keep_all = TRUE)
view(zip_df_unique)
```



##clean repeat Zips
all dates start with X so filter
```{r}
zori_df =
  read.csv(
    "./zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |>
  janitor::clean_names() 
  view(zori_df)
```

```{r}
zori_tidy = zori_df |> 
  # Pivot to long format
  pivot_longer(
    cols = starts_with("x"),
    names_to = "date",
    values_to = "zori") |> 
  mutate(
    date = str_remove(date, "x"),
  ) |> 
    separate(date, into = c("year", "month", "day"), sep = "_") |> 
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    month = month.name[month]) |>
  mutate(
    county_name = recode(
      county_name,
    "Richmond County" = "Staten Island",
    "New York County" = "Manhattan",
    "Kings County" = "Brooklyn",
    "Queens County" = "Queens",
    "Bronx County" = "Bronx")) |> 
  rename(borough = county_name, zip_code = region_name, zori_year = year, zori_month = month) |>
  select(-day)

```
view df to check that pivot worked accordingly
```{r}
view(zori_tidy)
```

join zori dataframe with zip dataframe on zip_code
```{r}
zori_zip_join = zori_tidy|> 
  left_join(zip_df_unique, by = "zip_code") |> 
  select ( metro, borough.x, neighborhood, zip_code, zori_year, zori_month, zori,)

view(zori_zip_join)
```

# Joined data descriptors
```{r}
total_observations = nrow(zori_zip_join)
print(paste("Total observations:", total_observations))

unique_zips = n_distinct(zori_zip_join$zip_code)
print(paste("Unique ZIP codes:", unique_zips))

unique_neighborhoods = n_distinct(zori_zip_join$neighborhood)
print(paste("Unique neighborhoods:", unique_neighborhoods))
```
In the joined dataset, there are 17284 total observations, 149 unique zip codes and 43 unique neighborhoods


# Find the ZIP codes that are in zip_df_unique but not in zori_tidy
```{r}
missing_zips = zip_df_unique |>
  anti_join(zori_tidy, by = "zip_code")
view(missing_zips)
num_missing_zips <- nrow(missing_zips)
print(num_missing_zips)
```
There were 171 zip codes that were in the zip code dataset but not in the zori data set. This could be due to many reasons such as heavy
commercial/industrial or private use within these zip codes that prevent rental activity. For example, after a quick google search the zip code
10464 was City Island in the Bronx that is a very small island with what appears to be beaches, parks and small shops. It is very 
likely that the rental activity is very low here which is why this zip code does not have a zori score.  

# Covid comparison
```{r}
price_comparison_table = zori_zip_join |> 
  # Filter for January 2020 and January 2021 data
  filter(zori_year %in% c(2020, 2021), zori_month == "January") |> 
  select(zip_code, borough.x, neighborhood, zori, zori_year) |> 
  # Pivot to a wider format to compare prices
  pivot_wider(
    names_from = zori_year,
    values_from = zori,
    names_prefix = "zori_"
  ) |> 
  # Calculate the change in price
  mutate(zori_change = round(zori_2021 - zori_2020, digits = 0)) |> 
  arrange(zori_change) |> 
  # Select the top 10 rows
   slice_head(n = 10) |>
  select(zip_code, borough.x, neighborhood, zori_change)
  print(price_comparison_table)
```
All of the neighborhoods with the largest drop in ziro scores were located in Manhattan which was hit particularly hard by the COVID-19
pandemic. COVID-19 spreads more rapidly in high population density environments which likely drove fear into the Manhattan rental 
market causing rental prices to drop dramatically during the pandemic.




